---
title: "Data Visualization - Part 2"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 3
    #code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE,warning=FALSE, cache=TRUE)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

*Learning goals and steps:*

*1.  Customize histograms in ggplot2*

|        *-Add titles, subtitles, and captions*
        
*2.  Barplots in ggplot2*

|        *-Reorder factors, modify legend*
        
*3.  More practice with scatterplots and regression lines*

*4.  Animate your plots*

*5.  Create your own ggplot theme*

*6.  Challenge yourself!*

      
<br>

**Due:** Nothing this week

Following from our first tutorial on data visualization using ggplot2, we are now back for more ggplot2 practice and customization. Since no two figures are ever the same, the ability to customize your figures is key. The ultimate aim of this tutorial is to help you to make your figures even more beautiful and informative! 

As with part 1, there are several plots that you will need to save for your write-up.  They are indicated with the <mark style="color:blue">*blue text*</mark>.

Before starting: 1) get the three datasets and packages from last week uploaded into R, 2) setup your script and, 3) ANNOTATE it!

<br>

### 1. Customize histograms in ggplot2

Use the code from last week to generate the histogram of flower-pollinator asynchrony. 

If we want to manipulate the x-axis to display every 5 values from -30 through 30 you can run the following code <mark style="background-color:grey">breaks = c(seq(-30,30,5))</mark> within the <mark style="background-color:grey">scale_x_continuous()</mark> function of ggplot2. Do you understand what the <mark style="background-color:grey">seq()</mark> function does?  Look up the help documentation if you are unsure.

We can also specify the limits of the plot axes – add the layers (of code) below to your <mark style="background-color:grey">ggplot()</mark> code and you’ll be able to see that the limit of the y axis now extends to the value of 160.  I’ll leave it to you to decide where to add these layers.  Remember that layers override each other if they modify the same features of the plot.

```{r, eval=T, echo=F}
#import datasets
Mid_bloom<-read.csv("C:/Users/17163/Documents/DSEE/Labs/Apple pollinators/Mid_bloom.csv",header=T)
Climate<-read.csv("C:/Users/17163/Documents/DSEE/Labs/Apple pollinators/Climate.csv",header=T)
Pollinators<-read.csv("C:/Users/17163/Documents/DSEE/Labs/Apple pollinators/Pollinators.csv",header=T)

#calculate asynchrony

#join mid bloom date with pollinators
Pollinators <- left_join(Pollinators, Mid_bloom, by = c("Year"))

#calculate asynchrony and add to the dataset using mutate
Pollinators <- Pollinators %>%
  mutate(Asynchrony = CorrectedCollectionDay - Mid_bloom_day)

#subset only asynchrony <= 25
Pollinators <- filter(Pollinators, abs(Asynchrony) <= 25)
```

```{r, eval=F, echo=T}
    scale_x_continuous(breaks = c(seq(-30,30,5))) + 
    scale_y_continuous(limits = c(0,160)) +
```

<br>

### 1a. Add titles, subtitles, and captions

Now it’s time for us to add more information to our graphs, for example, the plot title, subtitle, and partial caption. This might not be particularly useful in this case, but here’s some guidance just in case you do require it in your own work.

```{r, eval=F, echo=T}
   labs(title = "Plant-pollinator asynchrony", 
         subtitle = "For Malus ssp. (apple) flowers and bee pollinators",
         caption = "Data from Bartomeus et al. 2013")
```

<br>

#### <mark style="background-color:#F9E59B">Control everything!</mark>

<mark style="background-color:#F9E59B">Remember, you can also add in </mark><mark style="background-color:grey">theme()</mark><mark style="background-color:#F9E59B"> elements to your plot, which let you customize even more aspects! We already introduced theme elements in our [previous tutorial](https://ourcodingclub.github.io/tutorials/datavis/index.html). Other things you can play around with are:</mark>

|    <mark style="background-color:#F9E59B">- italicize or bold the text with </mark><mark style="background-color:grey">face = 'italic'</mark><mark style="background-color:#F9E59B"> or </mark><mark style="background-color:grey">face = 'bold'</mark><mark style="background-color:#F9E59B"> respectively</mark>

|    <mark style="background-color:#F9E59B">- center the title using </mark><mark style="background-color:grey">hjust = 0.5</mark>

<br>

Add the following code to your theme layer to modify the title of the plot.

```{r, eval=F, echo=T}
theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold"))
```

We can also italicize portions of titles or axes labels using <mark style="background-color:grey">expression()</mark>. For example, we can italicize the genus name Malus by substituting the following code into the theme layer.

```{r, eval=F, echo=T}
subtitle = expression("For"~italic("Malus")~"ssp. (apple) flowers and bee pollinators")
```


```{r, eval=T, echo=F}
#add title, subtitle and caption
(asynchrony_hist <- ggplot(Pollinators, aes(x = Asynchrony)) +                
    geom_histogram(binwidth = 2, color = "#8B5A00", fill = "#CD8500") +    # Changing the binwidth and colors
    geom_vline(aes(xintercept = mean(Asynchrony)),                       # Adding a line for mean abundance
               color = "red", linetype = "dashed", size=1) +           # Changing the look of the line
    theme_bw() +                                                      # Changing the theme to get rid of the grey background
    ylab("Frequency\n") +                                                   # Changing the text of the y axis label
    xlab("\nPollinator asynchrony")  +                              # \n adds a blank line between axis and text
    scale_x_continuous(breaks = c(seq(-30,30,5))) + 
    scale_y_continuous(limits = c(0,160)) +
    labs(title = "Plant-pollinator asynchrony", 
         subtitle = expression("For"~italic("Malus")~"ssp. (apple) flowers and bee pollinators"),
         caption = "Data from Bartomeus et al. 2013") +
    theme(axis.title.x = element_text(size = 14, face = "plain"),       # x axis face="plain" is the default, you can change it to italic, bold, etc. 
          axis.title.y = element_text(size = 14, face = "plain"),       # y axis 
          panel.grid = element_blank(),                                 # Removing the grey grid lines
          plot.margin = unit(c(1,1,1,1), units = , "cm"),               # Putting a 1 cm margin around the plot
          plot.title = element_text(size = 14, hjust = 0.5, face = "bold"))) 
```


<br>

### 2. Barplot to examine whether asynchrony differs between families

A barplot is a useful alternative to a boxplot.  In fact, I would say that they are used more frequently in many fields (including ecology) than boxplots.

The first thing we need to do is calculate the mean asynchrony for each family.  You should be able to generate that code on your own using the <mark style="background-color:grey">dplyr</mark> functionality.

```{r, eval=T, echo=F}
# generate object of Family mean Asynchrony values
Pollinator_fam <- Pollinators %>%
  group_by(Family) %>%
  summarize(Asynchrony = mean(Asynchrony, na.rm=T))
```

```{r}
# Plotting the mean asynchrony for families
(Asynch_barplot <- ggplot(Pollinator_fam, aes(x = Family, y = Asynchrony, fill = Family)) +
    geom_bar(position = position_dodge(), stat = "identity") +
    theme_bw() +
    ylab("Asynchrony\n") +                             
    xlab("Family")  +
    theme(axis.text.x = element_text(size = 12, angle = 45, vjust = 1, hjust = 1), 
          axis.text.y = element_text(size = 12),
          axis.title = element_text(size = 14, face = "plain"),                      
          panel.grid = element_blank(),                                          
          plot.margin = unit(c(1,1,1,1), units = , "cm")))
```

This is a pretty nice plot. However, as biologists, we should hopefully notice something is missing from these plots… ERROR BARS.  In order to add error bars to the plot, we must first calculate them.  Then, we will add them to the plot using the <mark style="background-color:grey">geom_errorbar()</mark> function.  From looking at the code, do you understand how the error bars are added?  Can you explain it in your own words?

```{r}
#calculate standard error 
Pollinator_fam <- Pollinators %>%
  group_by(Family) %>%
  summarize(Asynchrony_mean = mean(Asynchrony,  na.rm=T),
            AsyncSD = sd(Asynchrony,  na.rm=T),
            AsyncN = n(),
            AsynchSE = AsyncSD/sqrt(AsyncN))

# Plotting the mean asynchrony for families with SE
(Asynch_barplot <- ggplot(Pollinator_fam, aes(x = Family, y = Asynchrony_mean, fill = Family)) +
    geom_bar(position = position_dodge(), stat = "identity") +
    geom_errorbar(aes(ymin = Asynchrony_mean - AsynchSE, ymax = Asynchrony_mean + AsynchSE), width = .3,  position = "identity") +
    theme_bw() +
    ylab("Asynchrony\n") +                             
    xlab("Family")  +
    theme(axis.text.x = element_text(size = 12, angle = 45, vjust = 1, hjust = 1),  
          axis.text.y = element_text(size = 12),
          axis.title = element_text(size = 14, face = "plain"),                      
          panel.grid = element_blank(),                                          
          plot.margin = unit(c(1,1,1,1), units = , "cm")))
```

<br>

### 2a. Reorder factors

Sometimes you need to reorder the levels of a variable that is or could be a factor (e.g., Family). We often want to do this so that we can plot values in a specific order.

If we wanted to have the family Colletinae come before Andreninae, we would first have to reorder the data in the dataframe. From this point, after reordering the data, <mark style="background-color:grey">ggplot</mark> will always plot Colletinae before Andreninae. 

```{r}
# make Family a factor
Pollinator_fam$Family <- as.factor(Pollinator_fam$Family)

# Reordering the data
Pollinator_fam$Family <- factor(Pollinator_fam$Family, 
                             levels = c("Colletinae", "Andreninae", "Apinae", "Halictinae", "Xylocopinae"),
                             labels = c("Colletinae", "Andreninae", "Apinae", "Halictinae", "Xylocopinae"))
```

If you re-run the barplot from above, the columns and legend should be reordered appropriately.

<br>

### 2b. Modify the legend

We can change the legend title to “Bee Family,” by specifying <mark style="background-color:grey">name = "Bee Family"</mark> in <mark style="background-color:grey">scale_fill_manual()</mark>. In some cases, we might not want to have a title for the legend at all, which you can do by specifying in <mark style="background-color:grey">scale_fill_manual</mark>, <mark style="background-color:grey">name = NULL</mark>.  There are a lot of other ways to modify the look of the legend within <mark style="background-color:grey">theme()</mark>.  Try out some of the modifications on your plot!

```{r, eval=F, echo=T}
 theme(legend.title = element_text(face = "bold"),
          legend.position = "bottom", 
          legend.box.background = element_rect(color = "grey", size = 0.3))
```

Let’s cover some more of the <mark style="background-color:grey">theme()</mark> elements we’ve used in the examples above:

|    • <mark style="background-color:grey">legend.title</mark> allows you to change the font size of the legend, or its formatting (e.g. bold).
    
|    • The <mark style="background-color:grey">legend.position</mark> can be defined with accepted positions such as <mark style="background-color:grey">"bottom"</mark>, but you can also modify the coordinate position with something like this <mark style="background-color:grey">legend.position = c(0.1, 0.8)</mark>. This is a neat trick in some cases, where you have lots of blank space within your plot itself and want to fine-tune the legend position.
    
|    • Finally, we’ve used <mark style="background-color:grey">legend.box.background = element_rect()</mark> to create a light grey rectangle that surrounds the legend. If you don’t want this, you can just remove that line of code.
<br>
Don’t forget that we can use the <mark style="background-color:grey">scale_...()</mark> functions to customize both the color code AND the legend at once. The <mark style="background-color:grey">scale_fill_manual(values = c("your-color-1", "your-color-2", ...))</mark> function lets you decide on custom color values for solid elements (bars, boxplots, ribbons, etc.), and its counterpart <mark style="background-color:grey">scale_color_manual()</mark> works exactly the same for line elements (points in a scatter plot, regression lines, box or column outlines, etc.). You need to make sure you put in as many colors as there are factor levels in your data.

<br>

#### <mark style="background-color:lightblue">Need inspiration for your colors?</mark>

<mark style="background-color:lightblue">Remember, you can define colors using R’s [built-in colour names](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf) or by specifying their [Hex codes](https://www.color-hex.com/). The Colour Picker package is a great way to pick colors within the comfort of R Studio.</mark>

<br>

### 3. More practice with scatterplots and regression lines

In our first data visualization exercise, we added best fit lines to scatterplots.  Here, we will review this skill and expand upon it.

Using your amazing data wrangling skills, and a few new tricks we will add the climate data to the Mid_bloom dataset.  

The Bartomeus paper (from which these data come) used only April temperatures in their analysis.  To extract those values, we need to convert the date variable in the climate dataset from a Character class variable to a Date class variable.  Then, we will filter the April temperatures out, generate a new dataset with yearly average April temperatures (calculated from maximum daily values), and join it with the Mid_bloom dataset.

```{r, eval=F, echo=T}
# package for converting date
install.packages("lubridate")
```

```{r}
library(lubridate)

# convert "date" from chr to a Date class and specify current date format
Climate$DATE<- as.Date(Climate$DATE)

# calculate average April max temperature
April_temp<-filter(Climate,format.Date(DATE, "%m")== "04")

# create year variable
April_temp$Year<-year(April_temp$DATE)

# overwrite April_temp with yearly means
April_temp <- April_temp %>%
  group_by(Year) %>%
  summarize(Tmax = mean(TMAX, na.rm=T))

# join climate data with Mid_bloom data
Mid_bloom <- left_join(Mid_bloom, April_temp, by = c("Year"))
```

We can view this as a basic scatterplot in ggplot2:

```{r}
(MB_T_scatter <- ggplot(Mid_bloom, aes(x = Mid_bloom_day, y = Tmax)) +
  geom_point() +
  theme_bw())
```

We can see a pretty clear trend over time. So let’s try to plot a simple straight line through this using <mark style="background-color:grey">geom_smooth</mark> in ggplot2.  If you can’t remember how to do this, look up the code from last week!  <mark style="color:blue">*Beautify this plot and save it for your write-up.*</mark>

However, sometimes the relationship isn’t linear. An alternative would be to use a different equation. Do you remember the Keeling curve from our second week of lab?  You should reproduce that curve using ggplot2 (make it pretty!) and add a best fit straight line.  Then try a quadratic fit - something slightly more complicated to produce than the standard fits provided by R. Thankfully, ggplot2 lets us customize to pretty much any type of fit we want, as we can add in an equation to tell it what to plot. There are also several different base fits available. You can check out some here.

```{r, eval=F, echo=T}
# Modify your geom_smooth layer as follows to add a quadratic fit instead of a straight line fit
geom_smooth(method = "lm", formula = y ~ x + I(x^2)) +
```


Take some time to explore the different ggplot2 fits! For instance, <mark style="background-color:grey">method = "loess"</mark> gives a smoothed curve.  <mark style="color:blue">*Save your Keeling curve with the quadratic fit for your write-up.*</mark>

<br>

### 4. Animate your plots

Plotting best fit lines is a great way to emphasize a trend or pattern in a graph.  Another effective tool for visualizing patterns is to animate your plot.  We will animate the scatterplot you made of Mid bloom day by year using the <mark style="background-color:grey">gganimate</mark> package. 

The process sounds intimidating, but it is quite easy.  All you have to do is add a layer to your code telling R how you want to transition between states, in this case from year to year.  We use the <mark style="background-color:grey">transition_reveal()</mark> function for this.  Then you use a couple other functions that tell R to run the animation and save it as a .gif file.  Warning, the animation may take a minute to render and run in RStudio.  Be patient.

```{r, eval=F, echo=T}
# Get required packages
install.packages("gganimate")
install.packages("gifski")
```

```{r}
require(gganimate)
require(gifski)

# Plot mid bloom day by year
MB_scatter <- ggplot(Mid_bloom, aes (x = Year, y = Mid_bloom_day)) +
  geom_point(alpha = 0.5, size = 3) + 
  geom_line() + 
  theme_bw() +
  ylab("Mid bloom day\n") +                             
  xlab("\nYear")  +
  ylim(90,170) +
  theme(axis.text.x = element_text(size = 14, angle = 45, vjust = 1, hjust = 1),
        axis.text.y = element_text(size = 14),
        axis.title = element_text(size = 16, face = "plain"),                        
        panel.grid = element_blank(),                                                 
        plot.margin = unit(c(1,1,1,1), units = , "cm")) +                 
        transition_reveal(Year) # Reveals each new time frame
 
animate(MB_scatter, renderer = gifski_renderer()) # plays the animation in RStudio
anim_save("Mid_bloom_by_year.gif") # saves animation as a .gif to your working directory
```


Check out the [documentation](https://cran.r-project.org/web/packages/gganimate/gganimate.pdf) for the <mark style="background-color:grey">gganimate</mark> package.  There are a lot of different ways to animate plots besides just using the <mark style="background-color:grey">transition_reveal()</mark> function.

<br>

### 5. Create your own ggplot theme

You might have noticed that the lines starting with <mark style="background-color:grey">theme()</mark> quickly pile up. We’ve been adjusting the font size of the axes and the labels, the position of the title, the background color of the plot, etc. And then we’ve been copying and pasting those many lines of codes on all of our graphs, which really increases the length of our script, and makes our code less readable.

Here is a simple solution: create a customized theme that combines all the <mark style="background-color:grey">theme()</mark> elements you want! You can then apply it to your graphs to make things easier and increase consistency. You can include as many elements in your theme as you want, and when you apply your theme to a graph, only the relevant elements will be considered - e.g. for our histograms we won’t need to use <mark style="background-color:grey">legend.position</mark>, but it’s fine to keep it in the theme in case any future graphs we apply it to do have the need for legends.

```{r}
theme_coding <- function(){            # creating a new theme function
  theme_bw()+                          # using a predefined theme as a base
  theme(axis.text.x = element_text(size = 12, angle = 45, vjust = 1, hjust = 1),       # customizing lots of things
        axis.text.y = element_text(size = 12),
        axis.title = element_text(size = 14),
        panel.grid = element_blank(),
        plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), units = , "cm"),
        plot.title = element_text(size = 20, vjust = 1, hjust = 0.5),
        legend.text = element_text(size = 12, face = "italic"),
        legend.title = element_blank(),
        legend.position = c(0.9, 0.9))
}
```

Try out the effects of the theme by replacing all the code starting with <mark style="background-color:grey">theme(........)</mark> with just <mark style="background-color:grey">theme_coding()</mark>.

<br>

### 6. Challenge yourself!

<mark style="color:blue">*Continue flexing those new graphing muscles by making and saving the following plots for your write-up:*</mark> 

|       <mark style="color:blue">*1) A plot that examines the potential relationship between the climate and pollinator data.*</mark>

|       <mark style="color:blue">*2) A plot that is animated in some way to emphasize a meaningful pattern.*</mark>