---
title: "Introduction to R - Part 2"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 3
    #code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE,warning=FALSE, cache=TRUE)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

Learning goals:
    1.  Learn base R syntax for data manipulation
    2.  Turn messy data into tidy data with tidyr
    3.  Use efficient tools from the dplyr package to manipulate data
        ◦  Gain comfort with loops
Steps:
    1.  Subset, extract and modify data with R base operators
    2.  What is tidy data, and how do we achieve it?
    3.  Explore the most common and useful functions of dplyr
        ◦  Investigate for loops
    4.  Complete Introduction to R write-up
Due: 2/8/22 at 12:45 PM, Introduction to R write-up (20 pts.).  Includes everything asked of you in the blue questions within this document along with the additional materials and questions at the very end.  The write-up should be completed in your favorite word processor and uploaded to the assignment on Canvas.