---
title: "Introduction to R - Part 2"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 3
    #code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE,warning=FALSE, cache=TRUE)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy()
```

*Learning goals:*

*1.  Learn base R syntax for data manipulation*
  
*2.  Turn messy data into tidy data with tidyr*
  
*3.  Use efficient tools from the dplyr package to manipulate data*

|     *-Gain comfort with loops*
      
|  <br>    
      
*Steps:*

*1.  Subset, extract and modify data with R base operators*
  
*2.  What is tidy data, and how do we achieve it?*
  
*3.  Explore the most common and useful functions of dplyr*
  
|      *-Investigate for loops*
  
*4.  Complete Introduction to R write-up*
    
<br>

**Due:** 9/21/22 at 1:30 PM, Introduction to R write-up (20 pts.).  This assignment includes the plot from lab #1, everything asked of you in the <mark style="color:blue">*blue questions*</mark> within this document, along with the additional materials and questions at the very end.  The write-up should be completed in your favorite word processor and uploaded to the assignment on Canvas.

<br>

### Introduction

Data come in all sorts of different shapes and formats, and what is useful or practical for one application is not necessarily so for another. R has specific requirements about the setup and the types of data that can be passed to functions, so one of the best skills in your coding toolbox is being able to play with your data like putty and give it any shape you need!

We will start by showing some ways to manipulate data using base R syntax (without any extra package), because you will often see solutions online using this syntax, and it is good to understand how objects are built (and how to take them apart). After that, we will introduce principles of tidy data to encourage best practice in data collection and organization. We will then start using packages from the [Tidyverse](https://www.tidyverse.org/), a philosophy of coding that is quickly becoming the norm in R data science, and offers a potentially clearer way of coding than using only base R functions.

<br>

### 1.  Subset, extract, and modify data with R operators

Data frames are R objects made of rows and columns containing observations of different variables: you will often be importing your data that way. Sometimes, you might notice some mistakes after importing, need to rename a variable, or keep only a subset of the data that meets some conditions. Let’s dive right in and do that on the Mauna Loa data you worked with last week.

*Create a new, blank script, and add in some information at the top, for instance the title of the tutorial, your name, and the date (remember to use hashtags # to comment and annotate your script).*

*Also, add the code to download the data from the NOAA website (or the appropriate code for however you successfully accessed the data).  Don’t forget to name the variables too!*

We saw in part 1 that we can access variables in R by using the dollar sign <mark style="background-color:grey">$</mark>. This is already one way of subsetting, as it essentially reduces your data frame (2 dimensions) to a vector (1 dimension). You can also access parts of a data frame using square brackets <mark style="background-color:grey">[ , ]</mark>. The first number you put will get the row number, and the second the column. Leave one blank to keep all rows or all columns.

```{r, eval=T, echo=F}
MLO_CO2 <- read.table("https://gml.noaa.gov/aftp/data/trace_gases/co2/flask/surface/txt/co2_mlo_surface-flask_1_ccgg_month.txt") 
# This downloads data from the NOAA website.
```

```{r, eval=T, echo=F}
names(MLO_CO2) <- c("Site","Year","Month","CO2")   
# Applies names to each variable in the dataset
```


```{r}
# Here's how we get the value in the second row and fourth column
MLO_CO2[2,4]
```

Don’t remember what the fourth column is?  What function did we use in part one to print information about a data frame?  Reproduce that code in your script and run it!

```{r}
# Here's how we get all the info for row number 6
MLO_CO2[6, ]

# And of course you can mix it all together!
MLO_CO2[6, ]$CO2   # returns the value in the column CO2 for the sixth observation
# (much easier calling columns by their names than figuring out where they are!)
```

Subsetting with brackets using row and column numbers can be quite tedious if you have a large dataset and you don’t know where the observations you’re looking for are situated! And it’s never recommended anyway, because if you hard-code a number in your script and you add some rows later on, you might not be selecting the same observations anymore! That’s why we can use logical operations to access specific parts of the data that match our specification.

```{r}
# Let's access the values for the Year 1997
MLO_CO2[MLO_CO2$Year == 1997, ]
```

There’s a lot to unpack here! We’re saying: “Take this dataframe <mark style="background-color:grey">(MLO_CO2)</mark>, subset it (<mark style="background-color:grey">[ , ]</mark>) so as to keep the rows (writing the expression on the left-hand of the comma) for which the value in the column Year is exactly (<mark style="background-color:grey">==</mark>) 1997”. Note: The logical expression works here because the Year column contains numeric values: to access data that is of character or factor type, you would use quotation marks: <mark style="background-color:grey">MLO_CO2$Year == "1997"</mark>.

<br>

#### <mark style="background-color:lightblue">Operators for logical operations</mark>

<mark style="background-color:lightblue">Here are some of the most commonly used operators to manipulate data. When you use them to create a subsetting condition, R will evaluate the expression, and return only the observations for which the condition is met.</mark>

<mark style="background-color:grey">==</mark><mark style="background-color:lightblue">: equals exactly</mark>

<mark style="background-color:grey"><</mark><mark style="background-color:lightblue">, </mark><mark style="background-color:grey"><=</mark><mark style="background-color:lightblue">: is smaller than, is smaller than or equal to</mark>

<mark style="background-color:grey">></mark><mark style="background-color:lightblue">, </mark><mark style="background-color:grey">>=</mark><mark style="background-color:lightblue">: is bigger than, is bigger than or equal to</mark>

<mark style="background-color:grey">!=</mark><mark style="background-color:lightblue">: not equal to</mark>

<mark style="background-color:grey">%in%</mark><mark style="background-color:lightblue">: belongs to one of the following (usually followed by a vector of possible values)</mark>

<mark style="background-color:grey">&</mark><mark style="background-color:lightblue">: AND operator, allows you to chain two conditions which must both be met</mark>

<mark style="background-color:grey">|</mark><mark style="background-color:lightblue">: OR operator, to chains two conditions when at least one should be met</mark>

<mark style="background-color:grey">!</mark><mark style="background-color:lightblue">: NOT operator, to specify things that should be omitted</mark>

Let’s see them in action!

```{r}
# Subsetting with one condition
MLO_CO2[MLO_CO2$CO2 < 330, ]    # returns only the data for CO2 levels less than 320
MLO_CO2[MLO_CO2$CO2 <= 340, ]   # returns only the data for CO2 levels less or equal to 340

# This is completely equivalent to the last statement
MLO_CO2[!MLO_CO2$CO2 > 340, ]   # the ! means exclude

# Subsetting with two conditions
MLO_CO2[MLO_CO2$Year == 1997 | MLO_CO2$Year == 2010, ]    # returns only data for 1997 or 2010
MLO_CO2[MLO_CO2$Year == 1997 & MLO_CO2$Month %in% c(1:3), ]    # returns data for the first through third month of 1997
```

As you can see, the more demanding you are with your conditions, the more cluttered the code becomes. We will soon learn some functions that perform these actions in a more minimalist way, but sometimes you won’t be able to escape using base R (especially when dealing with non-data-frame objects), so it’s good to understand these notations.

<mark style="background-color:lightblue">Did you notice that last bit of code: </mark><mark style="background-color:grey">c(1:3)</mark><mark style="background-color:lightblue">? We saw in Part 1 that we can use </mark><mark style="background-color:grey">c()</mark><mark style="background-color:lightblue"> to concatenate elements in a vector. Using a colon between the two numbers means counting up from 1 to 3.</mark>

<mark style="background-color:lightblue">Other useful vector sequence builders are:</mark>

<mark style="background-color:grey">seq()</mark><mark style="background-color:lightblue"> to create a sequence, incrementing by any specified amount. E.g. try </mark><mark style="background-color:grey">seq(2, 12, 2)</mark>

<mark style="background-color:grey">rep()</mark><mark style="background-color:lightblue"> to create repetitions of elements. E.g. </mark><mark style="background-color:grey">rep(c(1,2), 3)</mark><mark style="background-color:lightblue"> will give </mark><mark style="background-color:grey">1 2 1 2 1 2</mark><mark style="background-color:lightblue">.</mark>

<mark style="background-color:lightblue">You can mix and match! What would </mark><mark style="background-color:grey">rep(seq(0, 30, 10), 4)</mark><mark style="background-color:lightblue"> give?</mark>

<mark style="background-color:lightblue">And finally, let’s say you need to modify some values or factor levels, or want to create a new column? Now that you know how to access parts of a data frame, you can do all of that. You only need an extra tool: the assign arrow </mark><mark style="background-color:grey"><-</mark><mark style="background-color:lightblue"> to overwrite data.</mark>

<br>

#### <mark style="background-color:#F9E59B">Creating and overwriting objects</mark>

<mark style="background-color:#F9E59B">Remember how we’ve been using the arrow <- to create new objects? This is a special convention in R that allows you to pick whichever name you want and assign it to an object (vector, list, data frame…).</mark>

<mark style="background-color:#F9E59B">Something to keep in mind is that if you use a name again in the same session, it will overwrite the former object. With experience, you can start making changes to an object and overwrite as you go, to “update” the object rather than creating many intermediaries (“object1”, “object2”, …). However, when you’re starting out, it’s a good idea to create these intermediary objects, or at least to create a “working copy” that you can reassign to the main data object once you’re satisfied with the changes.</mark>

<mark style="background-color:#F9E59B">As you will now see, we can also make use of the arrow </mark><mark style="background-color:grey"><-</mark><mark style="background-color:#F9E59B"> to overwrite specific values or range of values we need to change.</mark>

```{r}
## CHANGING VARIABLE NAMES AND VALUES IN A DATA FRAME

# Let's create a working copy of our object so we can practice making changes to our dataset without actually changing it
MLO_CO2b <- MLO_CO2

# Now suppose you want to change the name of a column: you can use the names() function
# Used on its own, it returns a vector of the names of the columns. Used on the left side of the assign arrow, it overwrites all or some of the names to value(s) of your choice.

names(MLO_CO2b)                 # returns the names of the columns

names(MLO_CO2b)[1] <- "site"    # Changing Site to site: we call the 1st element of the names vector using brackets, and assign it a new value
names(MLO_CO2b)[2] <- "Y"      # Changing Year to Y: we call the 2nd element and assign it the desired value

# Now suppose there's a mistake in the data, and the CO2 value 385.8 for March in 2008 should really be 386.8

## - option 1: you can use row and column number
MLO_CO2b[398,4] <- 386.8

## - option 2: you can use logical conditions for more control
MLO_CO2b[MLO_CO2b$Y == 2008 & MLO_CO2b$Month == 3, ]$CO2 <- 386.8   # completely equivalent to option 1
```

Can you spot pros and cons of options 1 and 2 above?

Option 1 is compact, but requires you to know exactly where the value to be corrected is. If you re-import your dataset at a later time with new values, it may not be in the same place.

Option 2 is longer and more difficult to read (it uses brackets to extracts the row corresponding to individual #373, and then the dollar sign to access just the column called X2008), but provides fine control, and the code will run even if the observation moves in your dataset.

<mark style="color:blue">*1.  Now it’s your turn to practice.  Write and run the code to change the CO2 value for May 2018 from 411.71 to 412 in the ML_CO2b dataset.  Include the code in your write-up.*</mark>

Using the same techniques, you can specify variable classes, which will be highly useful when we get to designing statistical models and need grouping variables like factors.

```{r}
## CREATING A FACTOR

# Let's check the classes
str(MLO_CO2b)

# The site column shows as character data, but it's really a grouping factor (the site could have been called A, B, C, etc.) Let's turn it into a factor:

MLO_CO2b$site <- as.factor(MLO_CO2b$site)        # converting and overwriting original class
str(MLO_CO2b)                                  # now site is a factor with 1 level
```

And what if you’re not happy with the factor levels? You can see the names of the factors with the <mark style="background-color:grey">levels()</mark> function… and yes, overwrite them, too.

That was a lot, but now you’ll be able to adapt these little chunks of code to manipulate your own data. The next sections will hopefully make things even easier, as they’ll teach you more intuitive functions to accomplish the same things.

<br>

### 2. What is tidy data, and how do we achieve it?

The way you record information in the field or in the lab is probably very different from the way you want your data entered into R. In the field, you want tables that you can ideally draw up ahead of time and fill in as you go, and you will be adding notes and all sorts of information in addition to the data you want to analyze. This usually results in data organized with many observations per row.  On the contrary, [tidy datasets](https://www.jstatsoft.org/article/view/v059i10) are arranged so that each row represents an observation and each column represents a variable. This is how our data is currently organized.

This makes a much longer dataframe row-wise, which is why this form is often called long format. Now if you wanted to compare between groups, treatments, etc., R would be able to split the dataframe correctly, as each grouping factor has its own column.

The <mark style="background-color:grey">spread()</mark> function from the <mark style="background-color:grey">tidyr</mark> package will let us convert long-format data (tidy dataframe) to wide-format. The <mark style="background-color:grey">gather()</mark> function will let us convert a wide-format table to a tidy dataframe. 

```{r,eval=F, echo=T}
install.packages("tidyr")  # install the package
```

```{r}
library(tidyr)             # load the package

# Spread() allows you to go from long to wide format
MLO_CO2_wide <- spread(MLO_CO2, Year, CO2)		          # in this order: data frame, key, value

# the key variable (Year) dictates the new columns (one for each year) and the value variable (CO2) indicates that CO2 concentrations are given within each Year column
#CO2 values are spread by year

# Let's reverse! gather() is the inverse function, allowing you to go from wide to long format
MLO_CO2_long <- gather(MLO_CO2_wide,Year, CO2,        # in this order: data frame, key, value
                                      c(3:50))        # we gather CO2 values by year (the first two of the 50 columns in MLO_CO2_wide are not CO2 values)
```

Since our original data was in long, tidy form, the new data frame MLO_CO2_long should be very similar to the original dataset.  However it is slightly different.  It added a few extra rows for the missing months of CO2 concentration in 1969 and 1976.

However, these functions have limitations and will not work on every data structure. To quote [Hadley Wickham](https://www.jstatsoft.org/article/view/v059i10), “every messy dataset is messy in its own way”. This is why giving a bit of thought to your dataset structure before doing your digital entry can spare you a lot of frustration later!

<br>

### 3. Explore the most common and useful functions of <mark style="background-color:grey">dplyr</mark>

The package <mark style="background-color:grey">dplyr</mark> is a fantastic bundle of intuitive functions for data manipulation, named after the action they perform. A big advantage of these functions is that they take your data frame as a first argument, so that you can refer to columns without explicitly having to refer to the full object (so you can drop those <mark style="background-color:grey">$</mark> signs!).  This is also easing us into the idea of piping, which is another aspect of the tidyverse that makes code readable for newer coders. Let’s meet the most common and useful functions.  Install and load the package if you haven’t done so already.

```{r, eval=F, echo=T}
install.packages("dplyr")  # install the package
```

```{r}
library(dplyr)              # load the package
```

<br>

### 3a. <mark style="background-color:grey">rename()</mark> variables

This lets you change the name(s) of a column or columns. The first argument is the data frame, the second (and third, etc.) takes the form New name = Old name.

```{r}
MLO_CO2_long <- rename(MLO_CO2_long, S = Site, M = Month, Y = Year, CO2 = CO2)     # changes the names of the columns and overwriting our data frame

# As we saw earlier, the base R equivalent would have been
names(MLO_CO2_long) <- c("S", "M", "Y", "CO2")
```

<br>

### 3b. <mark style="background-color:grey">filter()</mark> rows and <mark style="background-color:grey">select()</mark> columns

These are some of the most routine functions that let you reduce your data frame to just the rows and columns you need. The <mark style="background-color:grey">filter()</mark> function works great for subsetting rows with logical operations. The <mark style="background-color:grey">select()</mark> function lets you specify which columns to keep. Note: the <mark style="background-color:grey">select()</mark> function often clashes with functions of the same name in other packages, and for that reason it is recommended to always use the notation <mark style="background-color:grey">dplyr::select()</mark> when calling it.

```{r}
# FILTER OBSERVATIONS

# Let's keep observations from month 2 and 3 only, and from years 2009 to 2011

MLO_CO2_subset <- filter(MLO_CO2, Month %in% c(2, 3), Year %in% c(2009:2011)) # you can use multiple different conditions separated by commas

# For comparison, the base R equivalent would be (not assigned to an object here):
MLO_CO2[MLO_CO2$Month %in% c(2,3) & MLO_CO2$Year %in% c(2009:2011), ]
```

Note that here, we use <mark style="background-color:grey">%in%</mark> as a logical operator because we are looking to match a list of exact values. If you want to keep observations within a range of numeric values, you either need two logical statements in your <mark style="background-color:grey">filter()</mark> function, e.g. <mark style="background-color:grey">Month > 4 & Month <= 6</mark> or you can use the convenient <mark style="background-color:grey">between()</mark> function, e.g. <mark style="background-color:grey">between(Month, 4, 6)</mark>.

<mark style="color:blue">*2.  Write and run the code to create a new subset using the filter function that keeps values from months 8 through 12 in the years 1994 and 1995.  Include the code in your write-up.*</mark>

See how <mark style="background-color:grey">dplyr</mark> is already starting to shine by avoiding repetition and calling directly the column names without needing to call the object every time?

<mark style="background-color:#F9E59B">To quote or not to quote?</mark>

<mark style="background-color:#F9E59B">You may have noticed how we sometimes call values in quotes "", and sometimes not. This depends on:</mark>

<mark style="background-color:#F9E59B">- Whether the value you are calling is a character or numeric value: above, Month is of class integer (a number), so we don’t need quotes around the values it takes.</mark>
    
<mark style="background-color:#F9E59B">-Whether you are calling an existing object or referring to a value that R does not yet know about. Compare:</mark>
    
|        <mark style="background-color:grey">-new.object <- MLO_CO2</mark><mark style="background-color:#F9E59B"> and</mark>
        
|        <mark style="background-color:grey">-new.object <- "MLO_CO2"</mark>

<br>
        
<mark style="background-color:#F9E59B">The first creates a duplicate of our object, because R recognizes the name as an object in our environment. In the second case, you’re creating an object consisting of one character value.</mark>

<mark style="background-color:#F9E59B">It takes time and practice to get used to these conventions, but just keep an eye out for error messages and you’ll get there.</mark>

Now that we know how to subset rows, let’s do the same with columns!

```{r}
# SELECT COLUMNS

# Let's ditch the S column just as an example

MLO_CO2_no.S <- dplyr::select(MLO_CO2, Month, Year, CO2)   # or alternatively
MLO_CO2_no.S <- dplyr::select(MLO_CO2, -Site) # the minus sign removes the column

# For comparison, the base R equivalent would be (not assigned to an object here):
head(MLO_CO2[ , -1])  # removes first column
```

<br>

### 3c. <mark style="background-color:grey">group_by()</mark> certain factors to <mark style="background-color:grey">summarize()</mark> data

The <mark style="background-color:grey">group_by()</mark> function doesn’t cause visible change to your data frame. It creates an internal grouping structure. It’s very useful when you want to compute summary statistics using the <mark style="background-color:grey">summarize()</mark> function.

The <mark style="background-color:grey">summarize()</mark> function will always aggregate your original data frame, i.e. the output data frame will be shorter than the input. Here, let’s calculate the mean CO2 concentration (summarize) by year (group_by).

```{r}
# NEW OBJECT THAT SUMMARIZES OUR DATA USING THE GROUP_BY AND SUMMARIZE FUNCTIONS

MLO_CO2_y <- MLO_CO2 %>%                                 # specify data frame 
  group_by(Year) %>%		                                      # group by year
  summarize(CO2 = mean(CO2, na.rm=TRUE))	# calculate mean CO2 values (by year), na.rm=TRUE removes NA values
```

*This is our first true example of the piping method of coding*, which is more intuitive and readable than the base R functionality.

The <mark style="background-color:grey">%>%</mark> operator, called pipe, chains functions together.  So, the code above instructs R to use the MLO_CO2 data frame, group the data by Year, and calculate mean CO2 concentrations.  In other words, it is calculating mean CO2 values for ALL of the years in the dataset.

This is some really powerful code!  Remember how much coding you did last week just to plot the mean CO2 concentration for only 5 years.

Check out the structure of the new dataset and make a scatterplot of the CO2 concentrations by year. 

```{r}
str(MLO_CO2_y)

plot(CO2 ~ Year, data = MLO_CO2_y,
       ylab = "CO2 in parts per million (ppm)", xlab = "Year", ylim=c(300,420),
       cex.axis=1.5, cex.lab=1.5)
```

This plot should look familiar.  It is VERY similar to the [Keeling curve](https://keelingcurve.ucsd.edu/) we examined in class!

If you do not remember what some of these graphing parameters do, use <mark style="background-color:grey">help(“par”)</mark> to review them.

<mark style="color:blue">*3.  Use what you have just learned to create a data frame of mean CO2 concentrations by month (ignore year), and then plot them.  Play around with adjusting the graphing parameters to improve the plot aesthetics.  Include both plots of C02 by year and CO2 by month in your write-up along with the code used to make the plots.*</mark>

<br>

### 3d. <mark style="background-color:grey">..._join()</mark> datasets based on shared attributes

Sometimes you have multiple data files concerning the same project. Depending on your analytical needs, it may be very useful to have all the information in one table. This is where merging, or joining, datasets comes in handy.

The CO2 concentration data we have is interesting on its own.  However, it is not meaningful unless we can directly connect it to climate change.  Fortunately for us, NOAA also has temperature data for the Mauna Loa Observatory from 1977 to the present.  Unfortunately for us, the data is only available on the hourly scale and with each year in a separate text file.  This equates to 45 datasets each with 8760 rows.  We will need to do some serious wrangling in order to merge the temperature and CO2 data. 

We will once again use the <mark style="background-color:grey">read.table()</mark> function to download the text files from NOAA, but we will also use something called a <mark style="background-color:grey">for</mark> loop.  Loops are when a designated set of code is repeated, or run many times.  In a <mark style="background-color:grey">for</mark> loop, you specify the number of times the code is run (e.g., the number of years for which we have temperature data).  The other common type of loop is a <mark style="background-color:grey">while</mark> loop, where the code is repeated until a specified logical condition is satisfied (e.g., when year is equal to 2021, the last year of available data).

In the <mark style="background-color:grey">for</mark> loop below, you see that the number of iterations (repeats of code) is given immediately after <mark style="background-color:grey">for</mark> in parentheses.  Then, the code that is repeated is given in curly brackets <mark style="background-color:grey">{}</mark>.  In this case, the code is repeated 45 times.  Note, the code will take a bit of time to run.  Be patient.

Each repeat of the code downloads one years hourly temperature data, re-codes missing values, then calculates the mean (yearly) value.

Here is a link to the repository of climate data we used: https://gml.noaa.gov/aftp/data/meteorology/in-situ/mlo/

Within that repository is a [README](https://en.wikipedia.org/wiki/README) file that contains information about the repository, including the structure of the dataset: https://gml.noaa.gov/aftp/data/meteorology/in-situ/mlo/README

```{r}
# creating a data frame where yearly temperature means will be stored
MLO_temp <- data.frame(c(1977:2021),numeric(length(c(1977:2021))))
names(MLO_temp) <- c("Year","Temperature")

# for loop to extract temperature data from the NOAA website
# all of the code within the {} below will be run for each year (1977-2021) and the associated datasets on NOAA website
for (i in 1:nrow(MLO_temp)) {

  # reads in the yearly temperature datasets from NOAA
  MLO_t_data <- read.table(
    paste("https://gml.noaa.gov/aftp/data/meteorology/in-situ/mlo/met_mlo_insitu_1_obop_hour_",
          MLO_temp$Year[i],".txt",sep=""))	# can you figure out what the paste function does?
    
    # NOAA codes miss values as "-999.9", we convert those to NA
    MLO_t_data[MLO_t_data[,10]==-999.9,10]<-NA
    
   # extracts just the temperature data (column 10) and calculates the mean value
    MLO_temp$Temperature[i]<-mean(MLO_t_data[,10],na.rm=T)

}
```

Now that we have a temperature data frame.  The next step is to merge it with the yearly CO2 dataset.  We can do this because both datasets have a column representing the Year.

There are [many types of joins](https://dtplyr.tidyverse.org/reference/left_join.dtplyr_step.html) you can perform. They differ in how they handle data that is not shared by both tables, so always ask yourself which observations you need to keep and which you want to drop, and look up the help pages if necessary (in doubt, <mark style="background-color:grey">full_join()</mark> will keep everything). In the following example, we want to keep all the information in <mark style="background-color:grey">MLO_temp</mark> and <mark style="background-color:grey">left_join()</mark>.

```{r}
# Join the two data frames by Year. If the column names were spelled differently, we would need to tell the function which columns represent a match.

MLO_t_c <- left_join(MLO_temp, MLO_CO2_y, by = c("Year"))

# We see that the new object has the same length as our first data frame, which is what we want.
```

If the columns to match have the exact same name, you can omit them as they are usually automatically detected. However, it is good practice to specify the merging condition, as it ensures more control over the function.

Now that we have gone to the trouble of merging our data, let’s check if temperature and CO2 concentration are related to each other.

```{r}
# scatterplot of yearly temperature by CO2 concentration
plot(Temperature ~ CO2,data = MLO_t_c)
```

Are these variable related to each other? It is hard to tell without running an inferential statistical test.  We’ll go into more detail regarding how to test this in subsequent labs, but we can run a quick linear regression right now because who doesn’t like a little instant gratification.

```{r}
# linear regression of yearly temperature by CO2 concentration
# Values in the second row of the “Coefficients” table indicate the slope of the best fit line, SE, t-value, and p-value
summary(lm(Temperature ~ CO2, data = MLO_t_c))
```

<br>

### Assignment

<mark style="color:blue">*Your Introduction to R write-up should be completed in your favorite word processor, uploaded to the assignment on Canvas, and include:*</mark>

|    <mark style="color:blue">*A) The barplot from Introduction to R – part 1*</mark>
|    <mark style="color:blue">*B) The code and plots asked for in questions 1-4 of Introduction to R – part 2*</mark>
|    <mark style="color:blue">*C) Answers to the following questions:*</mark>
|        <mark style="color:blue">*1) What is one thing that you find useful about R?*</mark>
|        <mark style="color:blue">*2) What is one thing that you find frustrating?  What can you do (e.g., in coding practice or attitude, for example) to help limit that frustration?*</mark>
|        <mark style="color:blue">*3) There is a significant and positive linear relationship between Mauna Loa CO2 and temperature. However, the relationship has much more variation than the Keeling curve.  Given what we know about climate change, what is a possible explanation for the higher variation?*</mark>
|        <mark style="color:blue">*4) Are there any other interesting patterns or tid bits in the data that you would be interested in exploring further?  Why?*</mark>